---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "png",
  dev.args = list(type = "cairo-png"),
  fig.path = "figures/README_",
  out.width = "100%"
)
```

# INLAtools

<!-- badges: start -->
[![CRAN Status](http://www.r-pkg.org/badges/version-last-release/INLAtools)](https://cran.r-project.org/package=INLAtools)
[![](https://cranlogs.r-pkg.org/badges/INLAtools)](https://cran.r-project.org/package=INLAtools)
<!-- badges: end -->

Contain code to work with a C struct, in short cgeneric, to
    define a Gaussian Markov random (GMRF) model. The cgeneric contain
    required data to specify GMRF elements such as the graph and the
    precision matrix, as well hyper-parameters definition, such as 
    initial values and the prior, which are useful for model inference. 
    It can be accessed from a C program and is the
    recommended way to implement new GMRF models in the 'INLA' package
    (<https://www.r-inla.org>). We also provide access from R, 
    useful for code debug and evaluate the prior for multiple parameter sets.
    The implemented functionalities leverage the use of 'cgeneric' models.
    In order to illustrate it, a model named `generic0` is included.
    It can be used to implement intrinsic models and scale it by default,
    as proposed in Sørbye & Rue (2014) <doi:10.1016/j.spasta.2013.06.004>, 
    as well defining the required constraints.
    A very useful functionality is the Kronecker product
    method that creates a new model from multiple cgeneric models.
    It also works with the rgeneric, the R version of the cgeneric intended
    to easy try implementation of new GMRF models before making it for production.
    The constraints of a Kronecker between two cgeneric models, such as
    spatio-temporal intrinsic interaction models, see
  [Knorr-Held (2000)](https://onlinelibrary.wiley.com/doi/10.1002/1097-0258%2820000915/30%2919%3A17/18%3C2555%3A%3AAID-SIM587%3E3.0.CO%3B2-%23),
    are automatically set, as shown in the example 
    bellow.

## Installation

The 'INLA' package suggested for model fitting. 
You can install it with
```{r eval = !require(INLA), warning=FALSE, message=FALSE, comment=FALSE}
install.packages(
  pkgs = "INLA",
  repos = c(
    getOption("repos"),
    INLA="https://inla.r-inla-download.org/R/testing"), 
  dependencies = TRUE) 
```

You can install the current [CRAN](https://CRAN.R-project.org) version of INLAtools:
```{r cran-installation, eval = FALSE}
install.packages("INLAtools")
```

You can install the latest version of INLAtools from [GitHub](https://github.com/eliaskrainski/INLAtools) with
```{r ghinst, eval = FALSE}
if(!require(pak)) install.packages("pak")
pak::pkg_install("eliaskrainski/INLAtools")
```

## Example:

Build a Kronecker product model where the precision is 
given as 
\[ \mathbf{Q} = \tau \mathbf{R}_1 \otimes \mathbf{R}_2 \]

```{r first} 
library(Matrix)
## first
n1 <- 15
(R1 <- crossprod(diff(Diagonal(n1), differences = 2)))
```


```{r second}
G2 <- sparseMatrix(
  i = c(1,1,2,2,3,3,6), 
  j = c(2,3,4,6,4,5,7), 
  symmetric = TRUE)
(n2 <- nrow(G2)) 
R2 <- Diagonal(n = n2, x = colSums(G2)) - G2
R2
```

```{r viewK}
image(kronecker(R1, R2))
```

Notice that the order of the resulting matrix.

## Define the `cgeneric` models

We will use the 'cgeneric0' model for each. 
This model is written with precision matrix equal 
$\tau \mathbf{R}$ where the matrix $\mathbf{R}$ is the
precision structure matrix and $\tau$ is the precision parameter. 
A precision parameter multiplied by a precision structure is 
a local precision. 
```{r cg1}
library(INLAtools)
cg1 <- cgeneric(
    model = "generic0", 
    R = R1, ## precision structure matrix
    param = c(1, 0.05)) ## P(sigma > 1) = 0.05
```
where the prior is a PC-prior, as proposed in 
[Simpson et. al. (2017)](https://projecteuclid.org/journals/statistical-science/volume-32/issue-1/Penalising-Model-Component-Complexity--A-Principled-Practical-Approach-to/10.1214/16-STS576.full).

Notice that this model is intrinsically stationary,
so that using it as prior makes it an improper prior. 
The usual way to make it proper is to account for its
null space and define a constraint. 
It is done by default (`constr = TRUE`)
and the returned object already contain 
the 'extraconstr' to set it:
```{r cg1ctr}
str(cg1$f$extraconstr)
```

The implemented method to define such a constraint has 
complexity of $\mathcal{O}(n^3)$ because it perform the 
singular value decomposition. 
This is also used to scale the model.
This scaling, as proposed in 
[Sørbye & Rue (2014)](https://www.sciencedirect.com/science/article/pii/S2211675313000407),
is applied by default (`scale = TRUE`) and requires the 
generalized inverse of the $\textbf{R}$ matrix. 

If one know already what is the null space, 
it can be used in the `extraconstr` argument. 
Also, the `INLA` package provides a computationally efficient way 
to scale the model. 
Therefore, for large structure matrices, the better is to 
1) define a constraint with a known null space,
2) use it in `INLA::inla.scale.model()` to scale the model
3) use both to define the `cgneric` model. 
We will do this procedure for the second model.

For the second model we also use the `cgeneric0` but
fix the parameter to 1,
leaving the precision matrix from the first model
as the one for the resulting precision matrix.
Instead of using `scale = TRUE` and `constr = TRUE` (default), 
we consider an efficient (for larger models) way:
```{r cg2eff}
Ae2 <- list(A = matrix(1, 1, ncol(R2)), e=0) ## sum-to-zero constraint
R2scaled <- INLA:::inla.scale.model(Q = R2, constr = Ae2)
cg2 <- cgeneric(
  model = "generic0", 
  R = R2scaled, 
  param = c(1, NA), ## fix the precision/variance parameter
  extraconstr = Ae2,
  scale = FALSE, ### already scaled
  constr = FALSE ## already accounted
)
str(cg2$f$extraconstr)
```

The resulting Kronecker product model is
```{r cg12}
cg12 <- kronecker(cg1, cg2)
(N <- cg12$f$n)
```

Because each of the two model has one constraint, 
this new `cgeneric` model account for multiple constraints, 
the set of constraints in one model is replicated along each 
dimension of the other one. 
In the case of the resulting set of constraint contain 
redundant ones, these are removed.
For the our example, the final number of constraints is 
$n_2+n_1-1$:
```{r kconstr}
str(cg12$f$extraconstr)
```


```{r Q}
tau <- 4
Q <- cgeneric_Q(cg12, theta = log(tau))
image(Q)
```

Note: the scaling makes $\sigma = \tau^{-0.5}$
a marginal standard deviation parameter.

## Fit the model to some data

Simulate $k=5$ samples (replicates) from the 
Kronecker product model

```{r xx}
library(INLA)
k <- 5
xx <- inla.qsample(
  n = k, 
  Q = Q + Diagonal(N, 1e-9), 
  constr = cg12$f$extraconstr, 
  seed = 1
)
apply(xx, 2, summary)
```

Plot each replicate per group
```{r ggplot}
dataf <- data.frame(
  i1 = rep(rep(1:n1, each = n2), k),
  i2 = factor(rep(rep(1:n2, n1), k)),
  i = rep(1:N, k), 
  r = rep(1:k, each = N),
  x = as.vector(xx)
)
head(dataf, 10)

library(ggplot2)
ggplot(dataf) + theme_minimal() + 
  geom_line(aes(x = i1, y = x, group = i2, color = i2)) + 
  facet_wrap(~r)
```

Simulate observations considering a Poisson model as
```{r dataf}
dataf$y <- rpois(N * k, exp(3 + dataf$x))
```

Fit the model with a call to the main `INLA` function:
```{r inla}
fit <- inla(
    formula = y ~ f(i, model = cg12, replicate = r),
    family = "poisson", 
    data = dataf)
```

Summary of the intercept and $\tau$ posterior marginals
```{r smarg}
fit$summary.fixed
fit$summary.hyperpar
```

Scatterplot of the posterior mode and simulated
```{r sfit}
par(mar = c(3,3,0.5,0.5), mgp = c(2,0.5,0), bty = "n")
plot(fit$summary.random$i$mean, xx, pch = 8,
     xlab = "Posterior mean", ylab = "Simulated")
```

Posterior marginal summary for $\sigma=\tau^{-1/2}$:
```{r pmdsummary}
pm.sigma <- inla.tmarginal(
  fun = function(x) exp(-x/2), ## from log(tau) to sigma
  marginal = fit$internal.marginals.hyperpar[[1]])
1/sqrt(tau)
inla.zmarginal(pm.sigma)
```

