---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "png",
  dev.args = list(type = "cairo-png"),
  fig.path = "figures/README_",
  out.width = "100%"
)
```

# INLAtools

<!-- badges: start -->
[![CRAN Status](http://www.r-pkg.org/badges/version-last-release/INLAtools)](https://cran.r-project.org/package=INLAtools)
[![](https://cranlogs.r-pkg.org/badges/INLAtools)](https://cran.r-project.org/package=INLAtools)
<!-- badges: end -->

Contain code to work with a C struct, in short
  cgeneric, to define a Gaussian Markov random (GMRF) model.
  The cgeneric contain code to specify GMRF elements such as
  the graph and the precision matrix, and also the initial and 
  prior for its parameters, useful for model inference. 
  It can be accessed from a C program and is the recommended 
  way to implement new GMRF models in the 'INLA' package 
  (<https://www.r-inla.org>).
  The 'INLAtools' implement functions to evaluate 
  each one of the model specifications from R.
  The implemented functionalities leverage the use 
  of 'cgeneric' models and provide a way to debug 
  the code as well to work with the prior for the 
  model parameters and to sample from it. 
  A very useful functionality is the Kronecker product method 
  that creates a new model from multiple cgeneric models.
  It also works with the rgeneric, the R version of the 
  cgeneric intended to easy try implementation of new GMRF models.
  The Kronecker between two cgeneric models can be used to build the 
  spatio-temporal intrinsic interaction models
  [Knorr-Held (2000)](https://onlinelibrary.wiley.com/doi/10.1002/1097-0258%2820000915/30%2919%3A17/18%3C2555%3A%3AAID-SIM587%3E3.0.CO%3B2-%23)
  for what the needed constraints are automatically set
  as in the example below.

## Installation

The 'INLA' package is a suggested one, but you will need it 
for actually fitting a model. You can install it with
```{r eval = !require(INLA)}
install.packages("INLA",repos=c(getOption("repos"),INLA="https://inla.r-inla-download.org/R/testing"), dep=TRUE) 
```


You can install the current [CRAN](https://CRAN.R-project.org) version of INLAtools:
```{r cran-installation, eval = FALSE}
install.packages("INLAtools")
```

You can install the latest version of INLAtools from [GitHub](https://github.com/eliaskrainski/INLAtools) with
```{r ghinst, eval = FALSE}
## install.packages("remotes")
remotes::install_github("eliaskrainski/INLAtools")
```

## Example:

Build a Kronecker product model where the precision is 
given as 
\[ \mathbf{Q} = \tau \mathbf{R}_1 \otimes \mathbf{R}_2 \]

```{r first} 
library(Matrix)
## first
n1 <- 15
(R1 <- crossprod(diff(Diagonal(n1))))
```


```{r second}
G2 <- sparseMatrix(
  i = c(1,1,2,2,3,3,6), 
  j = c(2,3,4,6,4,5,7), 
  symmetric = TRUE)
(n2 <- nrow(G2)) 
R2 <- Diagonal(n = n2, x = colSums(G2)) - G2
R2
```

```{r viewK}
image(kronecker(R1, R2))
```

Notice that the order of the resulting matrix.

## Define the `cgeneric` models

We will use the 'cgeneric0' model for each. 
This model is written with precision matrix equal 
$\tau \mathbf{R}$ where the matrix $\mathbf{R}$ is the
precision structure matrix and $\tau$ is the precision parameter. 
A precision parameter multiplied by a precision structure is 
a local precision. 
A scaling, as proposed in 
[Sørbye & Rue (2014)](https://www.sciencedirect.com/science/article/pii/S2211675313000407),
is applied by default in the following code
```{r cg1}
library(INLAtools)
cg1 <- cgeneric(
    model = "generic0", 
    R = R1, ## precision structure matrix
    param = c(1, 0.05)) ## P(sigma > 1) = 0.05
```
where the prior is a PC-prior, as proposed in 
[Simpson et. al. (2017)](https://projecteuclid.org/journals/statistical-science/volume-32/issue-1/Penalising-Model-Component-Complexity--A-Principled-Practical-Approach-to/10.1214/16-STS576.full).

For the second model we also use the `cgeneric0` but
fix the parameter to 1,
leaving the precision matrix from the first model
as the one for the resulting precision matrix.

```{r cg2}
cg2 <- cgeneric(
    model = "generic0", 
    R = R2, ## precision structure matrix
    param = c(1, NA)) ## fix sigma = 1
```

The resulting Kronecker product model is
```{r cg12}
cg12 <- kronecker(cg1, cg2)
(N <- cg12$f$n)
tau <- 4
Q <- prec(cg12, theta = log(tau))
image(Q)
```

Note: the scaling makes $\sigma = \tau^{-0.5}$
a marginal standard deviation parameter.

## Fit the model to some data

Simulate $k=5$ samples (replicates) from the 
Kronecker product model

```{r xx}
library(INLA)
k <- 5
xx <- inla.qsample(
  n = k, 
  Q = Q + Diagonal(N, 1e-9), 
  constr = cg12$f$extraconstr
)
apply(xx, 2, summary)
```

Plot each replicate per group
```{r ggplot}
dataf <- data.frame(
  i1 = rep(rep(1:n1, each = n2), k),
  i2 = factor(rep(rep(1:n2, n1), k)),
  i = rep(1:N, k), 
  r = rep(1:k, each = N),
  x = as.vector(xx)
)
head(dataf, 10)

library(ggplot2)
ggplot(dataf) + theme_minimal() + 
  geom_line(aes(x = i1, y = x, group = i2, color = i2)) + 
  facet_wrap(~r)
```

Simulate observations considering a Poisson model as
```{r dataf}
dataf$y <- rpois(N * k, exp(3 + dataf$x))
```

Fit the model with a call to the main `INLA` function:
```{r inla}
fit <- inla(
    formula = y ~ f(i, model = cg12, replicate = r),
    family = "poisson", 
    data = dataf)
```

Summary of the intercept and $\tau$ posterior marginals
```{r smarg}
fit$summary.fixed
fit$summary.hyperpar
```

Scatterplot of the posterior mode and simulated
```{r sfit}
par(mar = c(3,3,0.5,0.5), mgp = c(2,0.5,0), bty = "n")
plot(fit$summary.random$i$mean, xx, pch = 8,
     xlab = "Posterior mean", ylab = "Simulated")
```

Posterior marginal summary for $\sigma=\tau^{-1/2}$:
```{r pmdsummary}
pm.sigma <- inla.tmarginal(
  fun = function(x) exp(-x/2), ## from log(tau) to sigma
  marginal = fit$internal.marginals.hyperpar[[1]])
1/sqrt(tau)
inla.zmarginal(pm.sigma)
```

