---
title: "Kronecker product models with `cgeneric`"
author: "Elias T Krainski and Håvard Rue"
date: "`r format(Sys.time(), '%B, %Y')`"
output:
  - rmarkdown::html_vignette
  - rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Kronecker product models with `cgeneric`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "png",
  dev.args = list(type = "cairo-png"),
  fig.path = "figures/cgK_",
  out.width = "100%"
)
```

# Abstract

In this vignette we illustrate the use of the Kronecker method 
of `cgeneric` models to build interaction models. 
We illustrate as the easier way to work 
with interaction between intrinsic models because the 
needed constraints are automatically set. 
One can use this to implement the spacetime models propose by 
[@knorrheld2000stii](https://onlinelibrary.wiley.com/doi/10.1002/1097-0258%2820000915/30%2919%3A17/18%3C2555%3A%3AAID-SIM587%3E3.0.CO%3B2-%23).

## The model definition

The Kronecker product approach was proposed in @clayton1996glmm. 
It present the rules for deriving the distribution of a interaction 
between two main effects. 
The case when the interaction is between two 
random effects, each set over a different domain,
the precision (or covariance) for the resulting random effect 
is the Kronecker product model between 
the precision (or covariance) from each main effect.
The covariance of the interaction effect given as
\[ \Sigma = \sigma^2 \mathbf{C}_1 \otimes \mathbf{C}_2 \]
where $\sigma^2$ is a marginal variance parameter, 
$\textbf{C}_1$ is the correlation matrix over the 
first domain and $\textbf{C}_2$ is the correlation matrix 
over the second domain. 

Similarly, the precision matrix is given as
\[ \mathbf{Q} = \tau \mathbf{R}_1 \otimes \mathbf{R}_2 \]
where the matrices $\mathbf{R}_1$ and $\mathbf{R}_2$ are 
structure matrices for random effects over each domain.
Thus, this way of building random interaction models 
result in a covariance/precision separable interaction.
Notice that we have one precision parameter $\tau$.

This idea was applied to build spatio-temporal models by 
@knorrheld2000stii.
Then, there are words like
separable, inseparable, non-separable associated with. 
In @knorrheld2000stii models for 
inseparable disease risk were proposed.
Clearly, the **disase risk** is what is spacetime inseparable.
The precision (or covariance) of such interaction effects 
**is separable** as it is computed as a Kronecker product between 
the precision (or covariance) matrices from each random effect domain.

Let us define the first model considering a structure matrix 
from a second order random walk model, a common model used for 
one dimensional domain such as time series, considering it with 
size $n_1$. 

```{r first} 
library(Matrix)
n1 <- 15
(R1 <- crossprod(diff(Diagonal(n1), differences = 2)))
```

Let us now define a second structure matrix for a second domain, 
that could, in the spacetime cases, represent 
the structure for a model over the spatial (discrete) domain.

```{r second}
G2 <- sparseMatrix(
  i = c(1,1,2,2,3,3,6), 
  j = c(2,3,4,6,4,5,7), 
  symmetric = TRUE)
(n2 <- nrow(G2)) 
R2 <- Diagonal(n = n2, x = colSums(G2)) - G2
R2
```

```{r viewK}
image(kronecker(R1, R2))
```

Notice that the order of the resulting matrix, 
considering the first domain first, generate a
structure matrix where the overall pattern is like 
the first structure matrix and the first $n_2$ arrow/colums 
pattern is given by the second structure matrix.

## The `cgeneric` models

The models we have considered are intrinsically stationary. 
It is easier to see that by in the first order random walk model 
where $n-1$ consecutive differences are used to model $n$ variables.
Here we consider the class of models where the precision is just 
a precision parameter times a structure matrix,
$\tau \mathbf{R}$ where the matrix $\mathbf{R}$ is the
precision structure matrix and $\tau$ is the precision parameter. 
This kind of models is available in INLA as a latent model 
named 'generic0'.
We have implemented this same model using the `cgeneric` interface, 
because then we can make combinations as illustrated in this paper. 
Please see the INLA vignette about the `cgeneric` models for details.
We called 'cgeneric0' and it can be defined from a
given structure matrix and the prior for the precision parameter.

A precision parameter multiplying a precision structure matrix is 
a local precision. 
Therefore, one need to do a scaling, as proposed in 
[@sorbye2014scaling](https://www.sciencedirect.com/science/article/pii/S2211675313000407),
in order to have a marginal precision parameter.
We do this by default in the following code
```{r cg1}
library(INLAtools)
cg1 <- cgeneric(
    model = "generic0", 
    R = R1, ## precision structure matrix
    param = c(1, 0.05)) ## P(sigma > 1) = 0.05
```
where the prior is a PC-prior, as proposed in 
[@simpson2017pcprior](https://projecteuclid.org/journals/statistical-science/volume-32/issue-1/Penalising-Model-Component-Complexity--A-Principled-Practical-Approach-to/10.1214/16-STS576.full).
This PC-prior is set from the probability statement
$P(\sigma > \sigma_0) = p$, 
for a given $\sigma_0$ and $p$.

The rank deficiency of the structure matrix is 2, 
as it is a structure matrix for a second random walk.
Then we have to impose two constraints, 
set by default and returned as `extraconstr`:
```{r c1}
nrow(cg1$f$extraconstr$A)
```

For the second model we also use the `cgeneric0` but
fix its precision parameter to 1,
leaving the precision matrix from the first model
as the one for the resulting precision matrix.
```{r cg2}
cg2 <- cgeneric(
    model = "generic0", 
    R = R2, ## precision structure matrix
    param = c(1, NA)) ## fix sigma = 1
```
where we set the probability for the prior assignment as NA
which is internally interpreted so that the precision is fixed. 
Again, by since by default the model is scaled, 
the precision is marginal and is the inverse of the 
reference standard deviation squared.
The number of constraints for this model is 1:
```{r c2}
nrow(cg2$f$extraconstr$A)
```

The resulting Kronecker product model built with
```{r cg12}
cg12 <- kronecker(cg1, cg2)
nrow(cg12$f$extraconstr$A)
```
where the required automatic constraints are accounted for. 
This number is equal the rank deficiency of 
$\textbf{R} = \textbf{R}_1 \times \textbf{R}_2$.
The rank of $\textbf{R}$ is $(n_1-2)(n_2-1)$, giving
it rank deficiency equals to  $n_1+2n_2-2$.

The (default) scaling makes $\sigma = \tau^{-0.5}$
in the next code as a marginal standard deviation parameter.
```{r cg12q}
(N <- cg12$f$n)
tau <- 4
Q <- prec(cg12, theta = log(tau))
image(Q)
```

## Fit the model to some data

Simulate $k=5$ samples (replicates) from the 
Kronecker product model

```{r xx}
library(INLA)
k <- 5
xx <- inla.qsample(
  n = k, 
  Q = Q + Diagonal(N, 1e-9), 
  constr = cg12$f$extraconstr, 
  seed = 1
)
apply(xx, 2, summary)
```

Plot each replicate per group
```{r ggplot}
dataf <- data.frame(
  i1 = rep(rep(1:n1, each = n2), k),
  i2 = factor(rep(rep(1:n2, n1), k)),
  i = rep(1:N, k), 
  r = rep(1:k, each = N),
  x = as.vector(xx)
)
head(dataf, 10)

library(ggplot2)
ggplot(dataf) + theme_minimal() + 
  geom_line(aes(x = i1, y = x, group = i2, color = i2)) + 
  facet_wrap(~r)
```

Simulate observations considering a Poisson model as
```{r dataf}
dataf$y <- rpois(N * k, exp(3 + dataf$x))
```

Fit the model with a call to the main `INLA` function:
```{r inla}
fit <- inla(
    formula = y ~ f(i, model = cg12, replicate = r),
    family = "poisson", 
    data = dataf)
```

Summary of the intercept and $\tau$ posterior marginals
```{r smarg}
fit$summary.fixed
fit$summary.hyperpar
```

Scatterplot of the posterior mode and simulated
```{r sfit}
par(mar = c(3,3,0.5,0.5), mgp = c(2,0.5,0), bty = "n")
plot(fit$summary.random$i$mean, xx, pch = 8,
     xlab = "Posterior mean", ylab = "Simulated")
```

Posterior marginal summary for $\sigma=\tau^{-1/2}$:
```{r pmdsummary}
pm.sigma <- inla.tmarginal(
  fun = function(x) exp(-x/2), ## from log(tau) to sigma
  marginal = fit$internal.marginals.hyperpar[[1]])
1/sqrt(tau)
inla.zmarginal(pm.sigma)
```

# Reference
